Towards Performance Portability in Earth-System Modelling with GOcean.

### 2015-09-08 docs/papers/shallow_performance/PSEA-2015_review_response.txt

> Main:

> 1. The results can be broadly separated into i) comparison with the
> original 'shallow' benchmark and ii) Optimisation of the Psy
> layer. They are presented in this order but to my mind this seems like
> the wrong way round. I don't think that there is technically a 'right
> way round' as they are very separate, but I did expect to read about
> the optimisation first. Also while reading the document, it mentioned
> 'original and the (best) PsyKAl ...' on page 6, it left me wondering
> what 'best' is because I hadn't reached the bit about optimisation at
> that stage. Suggest to reorder.

We have split the Results section into three subsections and
re-ordered, as suggested.

> Other

> 2. space after 'PsyKal' and before ',' on second line of abstract.
Fixed.

> 3. para 2, page 2: should there be 'of' on line 3? in the
  programming of models?

We're talking here about programming models such as OpenMP, CUDA etc. rather
than about constructing scientific models.

> 4. para 4, page 3: is there a version number of shallow used in the tests?

GRAHAM.

> 5. last full para, page 3: x- and y-directions?
Fixed.

> 6. para 1, page 4: is it one third per section?
Re-worded.

> 7. page 5, second transformation: remove second 'as'
Fixed.

> 8. para 3, page 6: 64, 128 ... should these be squared also on the figure too.
Fixed.

> 9. para 4, page 6: in the statement 'for the first two bars...' I
  would say that this is also true for the third bar also. This is gnu
  on the IvyBridge. Is that understood?

> 10. para6, page 6: remove the before to.
> 11. para6, page 6: difference seems to be in italics.

> 12. para 2, page 8: could you reword this; "on Haswell loop fusion
 achieves more than 90% while on Ivy Bridge it is fully-fusing the
 first loop nest", I found it difficult to follow.

> 13. Figure 4 and labeling the transformations: The fact that you
  have labeled the run instances 0-10 means its difficult to compare
  the labels with the bars (for me at least). Rather than reading the
  shading, I found my self counting them and I kept getting mixed up
  because the sixth bar corresponds to transformation 5. Further, in
  the text on page 9, there is reference to bar 6 which I assume is
  transformation 6 but bar 7... I understand why its been laid out
  this way but I think the transformations/vanilla could be labeled
  1-11?

> 14. para 3, page 9: potentially change "operation of Figure 5 is
  fused" to "operation (lower loop in Figure 5) is fused."

> 15. pare 1, page 10: in the sentence; "remain constant for the
> duration of each trip of the time-stepping loop" I've never heard
> that terminology before, is that standard? I would expect to hear
> something like "remain constant for the duration of each iteration
> of the time-stepping loop"

> 16. the introduction of Figure 5 help to explain a bit more the
> detail of the applied transformations. This was good and it would
> have been nice to have had a bit more detail on the transformations
> from the start rather than disbursed through the text.

----------------------- REVIEW 2 ---------------------

> The paper presents a very detailed and rigorous analysis of 6 such
> combinations across 3 platforms and 4 compilers with overall promising
> results. However the description of the PSyKAI architecture and the
> responsibilities and interplay between the different layers does not
> become entirely clear. The authors present very little background and
> rationale as to how PSyKAI compares to similar approaches in the
> field. It should also be made very clear in the introduction that the
> PSyKAI layer needs to presently be written *manually*, since this
> makes the current approach effectively "hand-optimised code".

We have altered the text to make it clearer that in this work we are
manually optimising the PSy layer in order to learn what
transformations must be supported in our code generation system (which
will be the subject of a future paper).

> Clarification questions
> -----------------------

> Introduction

> 1) p2: In which way has the PSyKAI design been influenced by OP2? How does it compare?

RUPERT.

> 2) p2: It is unclear what is meant by "local fields" in the kernel
> context: what is the granularity? Does a kernel (as written by the
> scientist) operate on the entire partition? A colum? A vertical
> slice of the partition? A single point? For the PSyKAI layer to be
> automatically generated, the kernel interface would need to be be
> well defined.

The text has been re-worked slightly to emphasise that a Kernel, like
the Algorithm layer, has no concept of parallelism. The nature of the
field it operates on depends on the application.

> 3) p2: If the algorithm layer operates on a full field and the kernels
> on "local fields", is the PSyKAI layer responsible for the domain
> decomposition?

Yes it is. The text has been re-worked to emphasise this.

> 4) p3: Who implemented the "original" Shallow version benchmarked
> against and for which platform has it been optimised?

Paul Swarztrauber originally wrote the code but the header text
records modifications by at least two other people. The hardware that
they were optimising for is not known. A line has been added to this effect.

> 5) p3: What does "retaining the I/O capability" imply? Is bitwise
> reproducability of results a requirement?

It simply implies that the solution fields can be written to disk at a
user-defined interval. Text has been added to this effect. Bit-wise
reproducibility is not a requirement.

> 6) p4: it is unclear from Figure 1 and the description on the previous
> page whether invoke_tstep is a routine written by the scientist and
> 'transformed' or entirely the responsibility of PSyKAI. It the latter
> is the case, how can PSyKAI determine which kernels to call, in which
> order, which parameters to pass etc.?

> Methodology

> 7) p5: Some of the transformations (3, 4, 5, 7, 9, 10) seem to be very
> specific to the Shallow application. In which way do those represent
> optimisations 'independent of the science code' as claimed?

Of necessity, the code transformations required to recover performance
for a specific code are determined by the structure of that code. We
are not claiming that a general set of optimisations can be applied to
any given code. Instead, the PSyKAl approach separates the code which
must be optimised from the code describing the science. The
optimisations are restricted to the PSy layer and therefore do not
affect the scientific components of the code.

> 8) p5: Could transformations be encoded as rules in a reusable form?
> Presently every version of the PSyKAI layer is presumably a separate,
> independent implementation?

In this paper we are concerned with whether (and how) it is possible
to recover performance after applying the considerable restructuring
required to achieve the PSyKAl separation of concerns. To do this we
have manually written the PSy layer. As the reviewer correctly states,
this means we would have a separate implementation for each target
architecture; precisely the situation we are working to avoid. We are
therefore developing the PSyclone code-generation system to generate
and manipulate the PSy layer. This will be the subject of a future
paper.

> Conclusions

> 9) p11: As pointed out in the comments above it seems the PSyKAI
> layer would require more tightly specified interface to both the
> algorithmic and kernel layers to be suitable for automatic
> generation. It would be interesting to hear the authors' thoughts on
> this.

The reviewer is correct, however automatic generation is not the
subject of this paper. Suffice to say we have such interfaces and a
python-based code-generation system (PSyclone) that can generate correct
code. This wil be the subject of a future paper.

> 10) p11: How would these user-provided transformation recipes look like?

These recipes are implemented as python scripts that apply PSyclone
Transformations.

General comments / suggestions
------------------------------

> 1) There seems to be no preprint available yet for reference [2]

> 2) Figure 2/3: A better visual grouping of the bars by platform and
> compiler e.g. by using colour to differentiate compilers and hatching
> to different platforms would help with interpretation of the plots

We would love to use colour but the formatting guidlines say we can't.

> 3) Figure 4: The bars per cluster are quite hard to tell apart visually

### 2015-09-10 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Add Flather paper to bib and cite it

The external forcing includes surface wind stress, bottom friction,
and open boundary barotropic forcing. A lateral slip boundary
condition is applied along the coast lines. The open boundary
conditions can be set as a clipped or a Flather's radiation
condition (Flather, 1976). The bottom friction takes a semi-implicit
form for the sake of model stability. As done in the original version
of NEMO, a constant or Smagorinsky horizontal viscosity coefficient is
used for the horizontal viscosity term.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal
version with that of the original version of the code. A negative value 
indicates that the \psykal version is faster than the original.}
\label{FIG_slowdown_summary}
\end{figure}

>>>>

The external forcing includes surface wind stress, bottom friction,
and open boundary barotropic forcing. A lateral slip boundary
condition is applied along the coast lines. The open boundary
conditions can be set as a clipped or a Flather's radiation
condition~\cite{flather76}. The bottom friction takes a semi-implicit
form for the sake of model stability. As done in the original version
of NEMO, a constant or Smagorinsky horizontal viscosity coefficient is
used for the horizontal viscosity term.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal
version with that of the original version of the code. A negative value 
indicates that the \psykal version is slower than the original.}
\label{FIG_slowdown_summary}
\end{figure}

### 2015-09-10 docs/papers/parco2015_abstract/psyclone_parco2015.tex

\author[A]{\fnms{A. R.} \snm{Porter}%
\thanks{Corresponding Author: Andrew Porter, STFC Daresbury Laboratory, Warrington, WA4 4AD, UK; E-mail:
andrew.porter@stfc.ac.uk.}},
\author[A]{\fnms{R. W.} \snm{Ford}},
\author[A]{\fnms{M.} \snm{Ashworth}},
\author[A]{\fnms{G. D.} \snm{Riley}}
and
\author[B]{\fnms{M.} \snm{Modani}}
\runningauthor{A.~R.~Porter et al.}
\address[A]{STFC Daresbury Laboratory, Warrington, WA4 4AD, UK}
\address[B]{IBM}

\begin{abstract}
This paper presents a domain specific approach to performance
portability and code complexity reduction in finite element and finite
difference codes. This approach has been developed for the UK Met
Office's next generation atmospheric model which uses finite elements
on a quasi-unform grid, and has also been prototyped on two finite
difference Ocean model benchmarks, one of which is based on the NEMO
ocean model. We first outline our approach and then present results
for a range of compilers. These demonstrate that it is possible to
retain and potentially improve upon the performance of an existing
shallow water model while simultaneously reducing its code complexity.
\end{abstract}

@@

\subsection{The \psykal Approach}

The \psykal approach separates code into three layers, the Algorithm
layer, the PSy layer and the Kernel layer. Whilst this approach is
general we have applied it to Atmosphere and Ocean models written in
Fortran where domain decomposition is typically performed in the
latitude-longitude direction, leaving columns of elements on each
domain-decomposed partition.

The top layer, in terms of calling hierarchy, is the Algorithm
layer. This layer specifies the algorithm that thescientist would like
to perform (in terms of calls to kernel and infrastructure routines)
and logically operates on full fields. We say logically here as the
fields may be domain decomposed, however the algorithm layer is not
aware of this. It is the scientists responsibility to write this
algorithm layer.

@@

The PSy layer sits in-between the Algorithm and Kernel layers and its
functional role is to link the algorithm calls to the associated
kernel subroutines. As the Algorithm layer works on logically global
fields and Kernel layer works on local fields the PSy layer is
responsible for iterating over columns. It is also responsible for
including any required distributed memory operations, such as halo
swaps and reductions.

@@

As an example, consider the following ``traditional'' code fragment.
\begin{verbatim}
!$OMP PARALLEL DO ...
do i = 1, nlat
  do j = 1, nlon
      do k = 1, levels
      a(k,j,i) = ...
      b(k,j,i) = ...
    end do
end do
end do
!$END OMP PARALLEL DO ...
\end{verbatim}
In the PSyKAl approach this code would be split into algorithm layer\ldots
\begin{verbatim}
call psy_...(a,b,...)
\end{verbatim}
PSy layer\ldots
\begin{verbatim}
subroutine psy_...(a,b,...)
  !$OMP PARALLEL DO ...
  do i = 1, nlat
    do j = 1, nlon
      call kern1(a,...)
      call kern2(b,...)
    end do
  end do
  !$END OMP PARALLEL DO
end subroutine psy_...
\end{verbatim}
and kernel layer\ldots
\begin{verbatim}
subroutine kern1(a,...)
  do k = 1, levels
    a(k) = ...
  end do
end subroutine kern1
subroutine kern2(b,...)
  do k = 1, levels
    b(k) = ...
  end do
end subroutine kern2
\end{verbatim}

@@

Clearly the splitting of code into separate layers will have an effect
on performance. This overhead and how to get back to the performance
of the ``traditional'' code, and potentially improve on it, will be
discussed in the talk.

\subsection{The `shallow' Program}
For this work we have used a benchmarking program that solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  In describing
our work, we term the version of the Shallow program obtained at the
beginning of this project the `original' version.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal version of shallow retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal version of shallow, we made the kernels
as fine-grained as possible. In this case, this resulted in eight
distinct kernels, each of which operated on a single field at a single
point (since we have chosen to use point-wise kernels). With a little
bit of tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
routine then consists of applying each of the kernels to all of the
points on the model mesh requiring an update. In its basic,
unoptimised (`vanilla') form, this PSy-layer routine then contains a
doubly-nested loop around each kernel call, as indicated by the
pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
       CPU       & Gnu   & Intel       & Cray    & IBM     \\
\hline
Intel Haswell (E5-1620 v2)   & 4.8.2 & 14.0.0      &         & \\
Intel Ivy Bridge  (E5-2697 v2) & 4.9.1 & 14.0.1.106  & 8.3.3 & \\
IBM Power 8      &       &             &             & 14.1    \\
\hline
\end{tabular}
\end{table}

Before applying any code transformations, we benchmark the original
version of the code. We also benchmark the vanilla, unoptimised
version after it has been re-structured following the \psykal
approach. These two versions of the code effectively provide upper and
lower bounds, respectively, on the performance we expect to achieve.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal
version with that of the original version of the code. A negative value 
indicates that the \psykal version is faster than the original.}
\label{FIG_slowdown_summary}
\end{figure}

@@

We have investigated the application of the \psykal separation of
concerns approach to the domain of finite-difference ocean
models. This approach enables the computational science (performance)
related aspects of a computer model to be kept separate from the
natural (oceanographic) science aspects. In this work we have shown
that the \psykal approach has the potential to at least maintain (if
not improve) performance while improving code
structure/maintainability.

The application of code transformations to the middle/PSy layer is key
to the performance of the \psykal version of a code. In our approach
this problem is tackled using the PSyclone code-generation system
which will be described in a future publication.

>>>>

\author[A]{\fnms{A. R.} \snm{Porter}%
\thanks{Corresponding Author: Andrew Porter, STFC Daresbury Laboratory, Warrington, WA4 4AD, UK; E-mail:
andrew.porter@stfc.ac.uk.}},
\author[A]{\fnms{R. W.} \snm{Ford}},
\author[A]{\fnms{M.} \snm{Ashworth}},
\author[A]{\fnms{G. D.} \snm{Riley}}
and
\author[B]{\fnms{M.} \snm{Modani}}
\runningauthor{A.~R.~Porter et al.}
\address[A]{STFC Daresbury Laboratory, Warrington, WA4 4AD, UK}
\address[B]{IBM India}

\begin{abstract}
This paper presents a domain specific approach to performance
portability and the reduction of code complexity in finite-element and
finite-difference codes. This approach has been developed for the UK
Met Office's next generation atmospheric model which uses finite
elements on a quasi-unform grid, and has also been prototyped on two
finite difference Ocean model benchmarks, one of which is based on the
NEMO ocean model. We first outline our approach and then present
results for a range of compilers. These demonstrate that it is
possible to retain and potentially improve upon the performance of an
existing shallow water model while simultaneously reducing its code
complexity.
\end{abstract}

@@

\section{The \psykal Approach}

The \psykal approach separates code into three layers; the Algorithm
layer, the PSy layer and the Kernel layer. While this approach is
general, we have applied it to Atmosphere and Ocean models written in
Fortran where domain decomposition is typically performed in the
latitude-longitude direction, leaving columns of elements on each
domain-decomposed partition.

The top layer, in terms of calling hierarchy, is the Algorithm
layer. This layer specifies the algorithm that the scientist would like
to perform (in terms of calls to kernel and infrastructure routines)
and logically operates on full fields. We say logically here as the
fields may be domain decomposed, however the algorithm layer is not
aware of this. It is the scientist's responsibility to write this
algorithm layer.

@@

The PSy layer sits in-between the Algorithm and Kernel layers and its
functional role is to link the algorithm calls to the associated
kernel subroutines. As the Algorithm layer works on logically global
fields and Kernel layer works on local fields, the PSy layer is
responsible for iterating over columns. It is also responsible for
including any required distributed-memory operations, such as halo
swaps and reductions.

@@

As an example, consider the following ``traditional'' code fragment:
\begin{verbatim}
!$OMP PARALLEL DO ...
do i = 1, nlat
  do j = 1, nlon
      do k = 1, levels
      a(k,j,i) = ...
      b(k,j,i) = ...
    end do
end do
end do
!$END OMP PARALLEL DO ...
\end{verbatim}
In the \psykal approach this code would be split into the {A}lgorithm
layer\ldots
\begin{verbatim}
call psy_...(a,b,...)
\end{verbatim}
PSy layer\ldots
\begin{verbatim}
subroutine psy_...(a,b,...)
  !$OMP PARALLEL DO ...
  do i = 1, nlat
    do j = 1, nlon
      call kern1(a,...)
      call kern2(b,...)
    end do
  end do
  !$END OMP PARALLEL DO
end subroutine psy_...
\end{verbatim}
and {K}ernel layer\ldots
\begin{verbatim}
subroutine kern1(a,...)
  do k = 1, levels
    a(k) = ...
  end do
end subroutine kern1
subroutine kern2(b,...)
  do k = 1, levels
    b(k) = ...
  end do
end subroutine kern2
\end{verbatim}

@@

Clearly the splitting of code into separate layers will have an effect
on performance. This overhead and how to get back to the performance
of the ``traditional'' code, and potentially improve on it, will be
discussed in the remainder of this paper.

\section{The `shallow' Program}

For the work described here we have used a benchmarking program that
solves the shallow-water equations on a bi-periodic plane, following
the finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  In describing
our work, we term the version of the Shallow program obtained at the
beginning of this project the `original' version.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal version of Shallow retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal version of Shallow, we made the kernels
as fine-grained as possible. This resulted in eight
distinct kernels, each of which operated on a single field at a single
point (since we have chosen to use point-wise kernels). With a little
bit of tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
routine then consists of applying each of the kernels to all of the
points on the model mesh requiring an update. In its basic,
unoptimised (`vanilla') form, this PSy-layer routine then contains a
doubly-nested loop around each kernel call, as indicated by the
pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevent element. The absence of an entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
       CPU       & Gnu   & Intel       & Cray    & IBM     \\
\hline
Intel Haswell (E5-1620 v2)   & 4.8.2 & 14.0.0      &         & \\
Intel Ivy Bridge  (E5-2697 v2) & 4.9.1 & 14.0.1.106  & 8.3.3 & \\
IBM Power 8      &       &             &             & 14.1    \\
\hline
\end{tabular}
\end{table}

Before applying any code transformations, we benchmark the original
version of the code. We also benchmark the vanilla, unoptimised
version after it has been re-structured following the \psykal
approach. These two versions of the code effectively provide
approximate upper and lower bounds, respectively, on the performance
we expect to achieve.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=90mm]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal
version with that of the original version of the code. A negative value 
indicates that the \psykal version is faster than the original.}
\label{FIG_slowdown_summary}
\end{figure}

@@

We have investigated the application of the \psykal separation of
concerns approach to the domain of finite-difference ocean
models. This approach enables the computational science (performance)
related aspects of a computer model to be kept separate from the
natural (oceanographic) science aspects. In this work we have shown
that the \psykal approach has the potential to at least maintain (if
not improve) performance while significantly improving code
structure/maintainability.

The application of code transformations/optimisations to the
middle/PSy layer is key to the performance of the \psykal version of a
code. In our approach this problem is tackled using the PSyclone
code-generation system which will be described in a future
publication.
